---
title: "Modeling The Freshening Power of the Hop Part 1: Tidy and Transform"
output: html_notebook
---

# outline


#Wickham:  four activities of analysis:
*tidy
*transform
*visualize
*model


# dplyr: filter, select, mutate, summarise, arrange (and group_by)
## five verbs of dplyr (Wickham)
* filter:  keep rows matching criteria
* select:  pick columns by name
* arrange: reorder rows (aka sort)
* mutate:  add new variables 
* summarise: reduce variables to values
* group_by: (the 6th verb?)


#Intro
This based on "ujbc_a_1469081_sm5496.txt" (supplementary data from Jacob A. Kirkendall, Carter A. Mitchell & Lucas R. Chadwick (2018): The Freshening Power of Centennial Hops, Journal of the American Society of Brewing Chemists DOI: 10.1080/03610470.2018.1469081  https://doi.org/10.1080/03610470.2018.1469081)
Downloaded 12/31/18 from https://ndownloader.figshare.com/files/11921120 via https://www.tandfonline.com/doi/suppl/10.1080/03610470.2018.1469081?scroll=top



# intro text!
if you care about what this means in a brewing context, first read the manuscript (link above).  This is simply an attempt to develop a metric such that "hop freshening power" can be incorporated into models of the warm-dryhopping process.  


# Goals
* express the ABW increase associated with dryhopping in terms of amount of hops added


# Restart R and Load packages
keep it simple for this one, just dplyr. 
```{r}
.rs.restartR()  # restart R  

#data wrangling
#install.packages("dplyr")  # dplyr "data plyer"  
library(dplyr)

#formating tables
#library(xtable)
library(flextable)  # to make more compact tables
#library(magrittr)



#text processing
#library(stringi)
#library(tidytext)
#library(yarrr)

#graphics
#library(ggplot2)
#library(ggvis)

#others
#library(qcc)
#library(lubridate) # lubridate for dates/times in general
#library(chron)      #  chron for converting fractional days to AM/PM/24HR/etc time formats


```



## csv import
```{r}

rm(list = ls()) # clear workspace
## 1 read.csv
mydata <-read.csv("ujbc_a_1469081_sm5496.txt",stringsAsFactors = FALSE)  ## SPECIFY filename

## 2 inspect imported data
```



#To do
*convert dates from 'character' to 'POSIX' date format
*calculate mean ABW increase relative to control (NH) for each experimental group.
**close inspection of the data shows that the beer in expt2 was not entirely end-fermented when dry-hopped.  This is evidenced by the fact that the control (NH) samples in those experiments increased by an average of ~0.15% ABV over the 42-day time-course of those experiments.  To account for this we will need to  create factor that allows us to group_by these discrete experimental groups (including separate groups for each 'expt2' time-point) so we can compute the mean baseline (NH) for each group.  
## reword the above!




## create new variables (*.POSIXct, timeofaddition, daysonhops, hops_g_100mL, pounds_bbl, variety, harvestyear,ox)
```{r}
### note: this is a base R data wrangling exercise! in general it's best to use lubridate package for date/timestamps! 

## can do one at a time (here we're creating new columns in POSIXct format, based on the particular date format used in the source data (ujbc_a_1469081_sm5496.txt; see ?as.POSIXct):
#x_brew_date.POSIXct<- as.POSIXct(mydata$brew_date, format = "%m/%d/%Y") # as a standalone vector (USELESS here), or:
#mydata$brew_date.POSIXct <-as.POSIXct(mydata$brew_date, format = "%m/%d/%Y")  # as a "new column" in our dataframe
  
## or, we can take advantage of the pattern that of our (character) date/time columns have the string "date" in the headername.  First make character vector of all column names containing string "date":
datecols<- dput(names(select(mydata, matches("date"))))  ## headers containing string "date"

## FORLOOP
# this forloop will (attempt to) convert datecols (define above) into POSIXct format (R datetime format).  In general they say it's  ### best to avoid forloops ### use functions from lubridate/etc packages with specific tools for the task at hand)!! ###
for (icol in datecols) {
  newcol = paste0(icol,".POSIXct")
  print(newcol)
  mydata[, newcol] = as.POSIXct(mydata[, icol],format = "%m/%d/%Y") ###  CREATE NEW columns with POSIXct   
#  mydata[, newcol] = as.POSIXct(as.numeric(mydata[, icol])  * (60*60*24), origin="1899-12-30") ###  microsoft times
}

## create timeofaddition, daysonhops, hops_g_100mL, pounds_bbl variables with dplyr "mutate"
mydata<- mydata %>% 
  mutate(timeofaddition = as.numeric(difftime(dryhop_date.POSIXct,brew_date.POSIXct)),
         daysonhops = as.numeric(difftime(Test_Date.POSIXct,dryhop_date.POSIXct)),
         hops_g_100mL = (mg_hops/volume_mL)/10,
         pounds_bbl = (mg_hops/volume_mL)*117/454
         )

mydata$OX<-grepl("OX", mydata$Hop_type)           ##  create logical "OX" column
mydata$Grind<-grepl("Grind", mydata$Hop_type)     ##  create logical column
mydata$Cone<-grepl("Cone", mydata$Hop_type)       ##  create logical column
mydata$harvest2014<-grepl("14", mydata$Hop_type)  ##  create logical column
mydata$harvest2015<-grepl("15", mydata$Hop_type)  ##  create logical column
mydata$harvest2017<-grepl("17", mydata$Hop_type)  ##  create logical column
## ifelse statement for harvestyear (if neither 2014 nor 2015 nor 2017, then 2016)
mydata$harvestYear <- ifelse(
  mydata$harvest2014==TRUE, 2014,
  ifelse(mydata$harvest2015==TRUE, 2015, 
         ifelse(mydata$harvest2017==TRUE, 2017, 2016)))
## ifelse statement for form of hops (if neither cone nor ground nor NH, then pellet) 
mydata$form_of_hops <- ifelse(
  mydata$Cone==TRUE, "cone",
  ifelse(mydata$Grind==TRUE, "ground",
         ifelse(mydata$Hop_type=="NH", "NH", "pellet")))

## ifelse statement for temperature greater or less than 10 
mydata$temperature_during_dryhop <- ifelse(
  mydata$temp.C<10, "cold",
  ifelse(mydata$temp.C>10, "warm", "something else"))

## create "variety" column starting with "Hop_type" then stripping away all the non-variety information
mydata$variety<-mydata$Hop_type
mydata$variety<- gsub("17","", mydata$variety)
mydata$variety<- gsub("16","", mydata$variety)
mydata$variety<- gsub("15","", mydata$variety)
mydata$variety<- gsub("14","", mydata$variety)
mydata$variety<- gsub("OX","", mydata$variety)
mydata$variety<- gsub("Grind","", mydata$variety)
mydata$variety<- gsub("Cone","", mydata$variety)
mydata$variety<- gsub(" ","", mydata$variety)

## create "EXPTnew" variable to account for beer not being end-fermented when time-series (expt 2A) commenced
mydata<- mydata %>% mutate(EXPTnew=paste0("group",expt, substr(special_group, 1,2)))
# clean it up by removing "NA" and any spaces due to canarycode bug
mydata$EXPTnew <- gsub(" ","", mydata$EXPTnew)  ## remove any spaces
mydata$EXPTnew <- gsub("NA","", mydata$EXPTnew) ## remove "NA"
```


Now we've created all variables we need to do our calculations.  First some cleanup:
```{r}
## rearrange columns (experimental factors on the left, then measurements, followed by calculations and finally all the date columns on the right):
mydata <- mydata %>% 
  select(sample_id,EXPTnew, hop, BINhop, variety, OX, harvestYear, form_of_hops, daysonhops, timeofaddition, temperature_during_dryhop, temp.C, hops_g_100mL, pounds_bbl,
         ABV, ABW, OE, Er, Ea, SG, RDF, ADF, Calories, 
         dhop_day, contact_days, REF_NH, ABV_increase, 
         brew_date.POSIXct, sample_collection_date.POSIXct, dryhop_date.POSIXct, Test_Date.POSIXct)
## remove ".POSIXct" suffix.  Leaving it as-is will only add to confusion if/when these data are saved and re-imported (and become 'character' format!)
colnames(mydata) = gsub(".POSIXct", "", colnames(mydata))
```



```{r}

#compute baseline ABW for each EXPTnew group
meanABW.REF_NH<- mydata %>% 
  group_by(EXPTnew) %>%
  filter(hop=="NH") %>%
  summarise(ABW.REF_NH=mean(ABW))
meanABW.REF_NH
```


## compute mean increase in ABW relative to meanABW.REF_NH (unhopped samples in same EXPTnew group), and normalize this relative to the exact amount of hops added
```{r}
## using objects created above...
## first join our data with mean ABV for unhopped samples in given experiment (meanABW.REF_NH; calculated above)
FPHcalc<- left_join(mydata, meanABW.REF_NH, by="EXPTnew")

## now calculate ABW_increase by subtracting each individual ABW measurement from meanABW.REF_NH:
FPHcalc$ABW_increase <- FPHcalc$ABW - FPHcalc$ABW.REF_NH

## and define "FPH" as the amount of ethanol produced per % dry-hops added (in g/100mL):
FPHcalc$FPH = FPHcalc$ABW_increase/FPHcalc$hops_g_100mL
FPHcalc <- FPHcalc %>% filter(hop=="DH")


#timezerorows
## since we know for certain that at time zero, the true dry-hop induced ethanol increase is zero.  Some time-zero rows we make "by hand" will reflect this reality much better than using real data that in this case.  (our real data at time zero is mostly instrument noise).
## we'll make three identical rows per


unique_rows <- !duplicated(FPHcalc[c("variety","OX","harvestYear", "form_of_hops", "temperature_during_dryhop")])  ## identify unique combinations
df <- FPHcalc[unique_rows,]     ## subset unique combinations
df$daysonhops<- 0
df$timeofaddition<- 0
df$ABV<- 0
df$ABW<- 0
df$OE<- 0
df$Er<- 0
df$Ea<- 0
df$SG<- 0
df$RDF<- 0
df$ADF<- 0
df$dhop_day<- 0
df$contact_days<- 0
df$ABV_increase<- 0
df$ABW.REF_NH<- 0
df$ABW_increase<- 0
df$FPH<- 0
timezerorows<-rbind(df,df)    ### duplicates
timezerorows<-rbind(timezerorows,df)  ### triplicates
#FPHcalc<-rbind(FPHcalc,timezerorows)  ### triplicates


## and save the transformed data to csv:
write.csv(FPHcalc,"FPHcalc.csv", row.names = FALSE)

##vanity table
tempster<- FPHcalc %>% group_by(variety,form_of_hops,daysonhops,harvestYear,OX, temperature_during_dryhop) %>%
  summarise_at(vars(hops_g_100mL,pounds_bbl, ABW_increase,FPH),funs(mean)) %>%
  arrange(desc(FPH)) 
tempster$OX<- as.character(tempster$OX)
tempster$daysonhops<- as.integer(tempster$daysonhops)
tempster$harvestYear<- as.integer(tempster$harvestYear)
tempster$ABW_increase<- substr(as.character(tempster$ABW_increase,2), 1,4)
tempster$hops_g_100mL<- substr(as.character(tempster$hops_g_100mL,2), 1,4)
tempster$pounds_bbl<- substr(as.character(tempster$pounds_bbl,2), 1,4)
tempster$FPH<- substr(as.character(tempster$FPH,2), 1,4)
tempster <- as.data.frame(tempster) 
tempster %>%  regulartable() %>% autofit()
```
We now have a single variable (FPH) that accounts for both the amount of hops added and the amount of ethanol produced.  We will use this variable to create models that [attempt to] relate variables such as hop variety, form of hops, time on hops, age and storage conditions of hops, and temperature during dryhopping, to the amount of ethanol produced due to dry-hopping.
# tranform complete.  We will use this output as basis for modeling work.

